An enhancer code compendium of early human neural and neural crest development

### 2025-04-25 manuscript.tex
Abstract

%Context
Enhancers encode cell types as clusters of transcription factor binding sites. 
%need
Enhancer code studies have been limited to specific cell types in model organisms.
%task and results
In this study we generate a single-cell multiome atlas of human neural tube organoids 
and of the head of a four weeks post conception human embryo. 
Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that 
are shared between the organoid model and the human embryo and we provide a compendium of enhancer codes 
for each individual cell type. 
We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a nucleosome depleted region. 
Enhancers encoding anterior-posterior patterning form an exception to this rule, being encoded as
homotypic clusters of homeodomain binding sites.
% TODO: S.A. [some more results? Validation examples? Genetic variants?]
%conclusion
This compendium of enhancer codes provides gene regulatory insight into human development.
%implication

>>>>

%Context
Enhancers encode cell types as clusters of transcription factor binding sites. 
%need
Enhancer code studies have been limited to specific cell types in model organisms.
%task and results
We generate a single-cell multiome atlas of human neural tube organoids 
and of the head of a four weeks post conception human embryo. 
Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that 
are shared between the organoid model and the human embryo and we provide a compendium of enhancer codes 
for each individual cell type. 
We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a nucleosome depleted region, using soft syntactic rules. 
We also provide counter examples to this rule, with enhancers encoded as homotypic clusters and by dimer motifs. 
Finally, we validate DeepNeuralTube using enhancer reporter assays and on human genetic variants.
%conclusion
The enhancer compendium is available as a resource at: something.aertslab.org
%implication
